---
title: "MovieLens Rating Prediction: A regularized  Bias Model with Iterative Coordinate Descent"
author: "Aariche aabderrahim"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: flatly
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4
)
```

```{r libraries, echo=FALSE}
library(tidyverse)
library(data.table)
library(caret)
library(lubridate)
```

```{r load-results, echo=FALSE}
# Load all pre-computed objects from the final model run
load("../RESULTS/movielens_results.RData")
```

------------------------------------------------------------------------

# Introduction & Overview

## Project Background

This report presents the methodology, implementation, and results of a
movie rating prediction system developed for the Harvard Data Science
Capstone. The objective is to build a model that predicts user ratings
on the MovieLens dataset with the lowest achievable Root Mean Squared
Error (RMSE).

The MovieLens dataset contains approximately 10 million ratings
submitted by users for a large catalogue of films. Each rating record
contains a user identifier, a movie identifier, a numerical rating (0.5
to 5.0 in 0.5 increments), a Unix timestamp, a movie title, and one or
more genre labels.

## Objective

The primary goal is to predict the rating $\hat{r}_{u,i}$ that user $u$
would assign to movie $i$, minimising the RMSE defined as:

$$\text{RMSE} = \sqrt{\frac{1}{N} \sum_{u,i} \left(\hat{r}_{u,i} - r_{u,i}\right)^2}$$

where $r_{u,i}$ is the true observed rating and $N$ is the total number
of predictions.

## Approach Summary

The model extends a regularized baseline approach — starting from the
global mean rating and sequentially estimating bias terms for movie,
user, year, month, week, and genre effects. Bias estimates are obtained
through iterative coordinate descent, where all effects are updated in
turn and the process repeats until convergence, producing stable and
robust estimates.

------------------------------------------------------------------------

# Methods & Model Description

## Data Partitioning

The `edx` dataset is split into a training set (90%) and a held-out test
set (10%) using stratified sampling via `createDataPartition`, with a
fixed seed for reproducibility.

Cold-start entries (users or movies appearing only in the test set) are
handled implicitly: missing bias values are imputed with zero, causing
predictions to fall back to the global mean or any available partial
bias.

## Feature Engineering

Three temporal features are extracted from the Unix timestamp using
`floor_date` (rather than `round_date`) to avoid attributing ratings to
future time periods. Genre labels, originally stored as pipe-delimited
strings, are parsed into individual genre tokens. A per-rating weight of
$1/n_\text{genres}$ is assigned so that a rating's total genre
contribution sums to 1 regardless of how many genres a movie belongs to.
A stable `row_id` is assigned immediately after feature engineering and
preserved throughout all subsequent operations to ensure correct joins.

## Baseline Model

The simplest possible predictor is the global mean rating $\mu$ computed
over the training set:

$$\hat{r}_{u,i} = \mu$$

This serves as the performance baseline against which all model
improvements are measured.

## Regularized Bias Model

The full prediction is formulated as a sum of the global mean and six
additive bias terms:

$$\hat{r}_{u,i} = \mu + b_i + b_u + b_y + b_m + b_w + b_g$$

where:

| Term | Description |
|----|----|
| $\mu$ | Global mean rating |
| $b_i$ | Movie bias — captures whether a film is rated above or below average |
| $b_u$ | User bias — captures individual rating tendencies |
| $b_y$ | Year bias — captures year‑level temporal trends |
| $b_m$ | Month bias — captures month‑level seasonal patterns |
| $b_w$ | Week bias — captures week‑level fluctuations |
| $b_g$ | Genre bias — captures genre preference patterns (weighted) |

Each bias term is estimated using regularized least squares. For a
generic effect $b_x$ with regularization parameter $\lambda$:

$$b_x = \frac{\sum_{j \in S_x} r_j^*}{|S_x| + \lambda}$$

where $r_j^*$ is the residual after removing all other known effects,
and $S_x$ is the set of ratings belonging to group $x$. Larger $\lambda$
shrinks estimates for groups with few observations toward zero, guarding
against overfitting on rare movies, users, or time bins.

The genre bias is computed at the level of individual genre tokens. Each
rating's genre contribution is the weighted sum of the bias values of
all genres it belongs to:

$$b_g^{(j)} = \sum_{g \in G_j} w_j \cdot b_g$$

where $w_j = 1/n_{\text{genres},j}$ and $G_j$ is the set of genres for
rating $j$. This is implemented via an unnested genre table joined back
to the rating level, avoiding `rowwise()` iteration entirely.

## regularization Parameter Tuning

The regularization parameter $\lambda$ is selected via a held-out
validation split of the training set (80% training, 20% validation). For
each candidate value in $\lambda \in \{1, 1.5, 2, \ldots, 10\}$, all six
bias terms are estimated on the training portion and evaluated on the
validation portion. The $\lambda$ minimising validation RMSE is retained
for final model training.

## Iterative Coordinate Descent

Rather than estimating each bias term in a single sequential pass, the
model employs iterative coordinate descent: all six bias terms are
updated in turn, each time conditioning on the current estimates of all
other terms. This process repeats until the change in training RMSE
falls below a convergence tolerance of $10^{-5}$ or a maximum of 30
iterations is reached.

At each iteration, all current bias values are joined onto the training
data to form a base table. Each effect is then updated by computing the
residual after removing the contributions of all other effects, and
applying the regularized sum formula. The genre effect is handled by
joining through the unnested genre table and re‑aggregating to the
rating level.

------------------------------------------------------------------------

# Code Engineering & Design Decisions

## Reusable Functions

Three operations are needed in multiple places — feature engineering,
genre contribution lookup, and full prediction assembly. Each is
encapsulated in a named function (`add_features()`,
`compute_genre_contrib()`, `build_predictions()`). This follows the DRY
principle and ensures consistency.

## Stable Row Identity

A unique, stable integer identifier `row_id` is stamped once at the
source (immediately after feature engineering) and never reassigned. It
serves as the join key that reconnects unnested genre rows back to their
parent rating. This prevents the latent bug of generating new row IDs
inside a function after the data has been reordered.

## Vectorised Genre Handling

Instead of using `rowwise()` loops (which are extremely slow on large
data), the script expands the genre list to one row per genre token,
performs vectorised joins, and then aggregates back to the rating level.
This uses optimised C‑level operations and runs orders of magnitude
faster.

## Pre‑allocation & Hoisting

Result containers are pre‑allocated with their final length to avoid
repeated memory reallocation. The expensive unnesting operation is
performed once outside the training loop and reused across all
iterations.

## NA Imputation for Cold‑Start

Missing bias values after `left_join` are replaced with zero, so that
unseen users, movies, or genres fall back to the global mean plus any
available partial information.

## Reproducibility

Explicit `set.seed()` calls with different seeds for the train/test
split and validation split ensure that all random steps are
reproducible. \## Model Serialisation: Reusable Predictor Closure

After training, the model is saved not only as raw bias tables but also
as a **closure** (a function that carries all bias values inside it).
The `make_predictor()` function captures the global mean $\mu$ and the
six bias tables, and returns a single function `predict_ratings()`.

This predictor function: - Takes a data frame with columns `userId`,
`movieId`, `timestamp`, `genres`. - Automatically performs feature
engineering (temporal binning, genre splitting, weighting). - Joins all
pre‑computed bias terms (coalescing missing values to zero). - Returns a
vector of predicted ratings.

The predictor is saved as an `.rds` file (`predict_ratings.rds`).
Loading it with `readRDS()` restores a fully functional prediction
function – no retraining, no repeated code, and no risk of inconsistent
feature engineering.

**Example usage:**

```{r predict_ratings, eval=FALSE,echo=TRUE}


predict_ratings <- readRDS("../RESULTS/predict_ratings.rds")

new_ratings <- data.frame(
  userId    = c(343, 233),
  movieId   = c(1123, 1123),
  title = c("Perfect Candidate, A (1996)","Perfect Candidate, A (1996)"),
  timestamp = c(974595468, 974595468),
  genres    = c("Documentary", "Documentary")
)

predictions <- predict_ratings(new_ratings)
new_ratings$p_rating <- round(predictions,2)
new_ratings$r_rating <- round(round(predictions, 1)/0.5)*0.5


knitr::kable(
  new_ratings,
  caption = "Predictions using the deployment‑ready predictor",
  col.names = c("userId", "movieId","title", "timestamp", "genres","p_rating", "r_rating"),
  align = c("l", "l", "c","c", "c", "r","r")
)

```

------------------------------------------------------------------------

# Results & RMSE

## Baseline Performance

The global mean predictor achieves an RMSE of
**`r round(rmse_baseline, 4)`** on the test set.

## Lambda Tuning Results

The optimal regularization parameter was found to be
**`r best_lambda`**, which minmizes the validation RMSE as shown below.

```{r lambda-plot}
plot(lambdas, val_rmse, type = "b",
     xlab = "Lambda", ylab = "Validation RMSE",
     main = "Lambda Tuning: Validation RMSE vs regularization Strength")
abline(v = best_lambda, col = "red", lty = 2)
legend("topright", legend = paste("Best lambda =", best_lambda),
       col = "red", lty = 2)
```

## Step-by-Step RMSE Improvement

The table below shows the RMSE on the test set as each bias term is
added incrementally, demonstrating the individual contribution of every
effect.

```{r step-rmse-table}
step_results <- data.frame(
  Step = c(
    "Baseline (global mean)",
    "+ Movie bias",
    "+ User bias",
    "+ Year bias",
    "+ Month bias",
    "+ Week bias",
    "+ Genre bias"
  ),
  RMSE = round(c(
    rmse_step_baseline,
    rmse_step_movie,
    rmse_step_user,
    rmse_step_year,
    rmse_step_month,
    rmse_step_week,
    rmse_step_genre
  ), 4),
  Improvement = round(c(
    0,
    rmse_step_baseline - rmse_step_movie,
    rmse_step_movie    - rmse_step_user,
    rmse_step_user     - rmse_step_year,
    rmse_step_year     - rmse_step_month,
    rmse_step_month    - rmse_step_week,
    rmse_step_week     - rmse_step_genre
  ), 4)
)

knitr::kable(
  step_results,
  caption  = "Test RMSE after adding each bias term",
  col.names = c("Model Step", "Test RMSE", "RMSE Reduction"),
  align    = c("l", "c", "c")
)
```

## Final Model Performance

After iterative coordinate descent training with the selected $\lambda$,
the model is evaluated on the held‑out test set. The final test RMSE is
**`r round(rmse_final, 4)`**.

## Interpreting the RMSE Table: Task‑Dependent Decisions

```         

The improvement column shows how much each bias term reduces RMSE *under a random split where all movies appear in both training and test sets*. However, the optimal model differs by deployment task. Use this decision map to avoid “RMSE delusion” – treating the numbers as absolute.
```

```{r Task‑dependent interpretation of the RMSE table }
decision_map <- data.frame(
  Task = c("**Static catalogue** (same movies, only new ratings)",
           "**Cold‑start** (new, unseen movies)"),
  What = c("Movie and user biases dominate; genre adds near‑zero.",
           "The table *underestimates* genre because movie bias steals its signal."),
  Action = c("Keep effects above a threshold (e.g., 0.0004). Drop genre bias – it adds no value. Model stays simple and fast.",
             "**Cannot rely on this table.** Build a separate evaluation: hold out entire movies. Then genre bias will likely show large improvement.")
)

knitr::kable(decision_map, caption = "Task‑dependent interpretation of the RMSE table")
```

## Summary Table

```{r results-table}
results <- data.frame(
  Model = c(
    "Baseline (global mean)",
    "regularized  Bias Model (Coordinate Descent)"
  ),
  RMSE = c(
    round(rmse_baseline, 4),
    round(rmse_final, 4)
  ),
  `Improvement (%)` = c(
    "—",
    round((rmse_baseline - rmse_final) / rmse_baseline * 100, 2)
  )
)

knitr::kable(
  results,
  caption   = "Model performance on the held-out test set",
  col.names = c("Model", "Test RMSE", "Improvement over Baseline (%)"),
  align     = c("l", "c", "c")
)
```

------------------------------------------------------------------------

# Visualizations & Plots

## Rating Distribution

```{r rating-dist, message=TRUE, warning=TRUE}
train_set |>
  ggplot(aes(x = rating)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", colour = "white") +
  scale_x_continuous(breaks = seq(0.5, 5, 0.5)) +
  labs(
    title = "Distribution of Ratings in the Training Set",
    x     = "Rating",
    y     = "Count"
  ) +
  theme_minimal()
```

## Ratings per Movie (Log Scale)

```{r ratings-per-movie}
train_set %>%
  count(movieId) %>%
  ggplot(aes(x = n)) +
  geom_histogram(bins = 50, fill = "steelblue", colour = "white") +
  scale_x_log10() +
  labs(
    title = "Number of Ratings per Movie (log scale)",
    x     = "Number of Ratings",
    y     = "Number of Movies"
  ) +
  theme_minimal()
```

## Average Rating Over Time (Year)

```{r avg-rating-year}
train_set %>%
  group_by(year_bin) %>%
  summarise(avg_rating = mean(rating), .groups = "drop") %>%
  ggplot(aes(x = year_bin, y = avg_rating)) +
  geom_line(colour = "steelblue") +
  geom_point(colour = "steelblue") +
  labs(
    title = "Average Rating by Year",
    x     = "Year",
    y     = "Average Rating"
  ) +
  theme_minimal()
```

## Top Movie Biases

```{r top-movie-biases}
if (exists("edx")) {
  movie_bias_final %>%
    left_join(edx %>% distinct(movieId, title), by = "movieId") %>%
    arrange(desc(b_i)) %>%
    slice_head(n = 15) %>%
    ggplot(aes(x = reorder(title, b_i), y = b_i)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    labs(
      title = "Top 15 Movies by Positive Bias (b_i)",
      x     = NULL,
      y     = "Movie Bias"
    ) +
    theme_minimal()
} else {
  cat("Movie titles not available – plot omitted.\n")
}
```

## Genre Bias

```{r genre-bias}
genre_bias_final %>%
  arrange(desc(b_g)) %>%
  ggplot(aes(x = reorder(genre_single, b_g), y = b_g)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Genre Bias Estimates",
    x     = "Genre",
    y     = "Genre Bias (b_g)"
  ) +
  theme_minimal()
```

## Residual Distribution

```{r residuals}
test_pred %>%
  mutate(residual = rating - pred) %>%
  ggplot(aes(x = residual)) +
  geom_histogram(binwidth = 0.1, fill = "steelblue", colour = "white") +
  labs(
    title = "Distribution of Prediction Residuals on Test Set",
    x     = "Residual (True − Predicted)",
    y     = "Count"
  ) +
  theme_minimal()
```

## Residuals vs Predicted (Diagnostic)

```{r residual-plot}
test_pred %>%
  mutate(residual = rating - pred) %>%
  ggplot(aes(x = pred, y = residual)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(color = "blue", se = FALSE) +
  labs(
    title = "Residuals vs. Predicted Ratings",
    x     = "Predicted rating",
    y     = "Residual"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

# Conclusion

## Summary

This project developed a regularized additive bias model for movie
rating prediction on the MovieLens dataset. Starting from a global mean
baseline (RMSE = `r round(rmse_baseline, 4)`), the final model trained
via iterative coordinate descent achieves a test RMSE of
`r round(rmse_final, 4)` — an improvement of
`r round((rmse_baseline - rmse_final)/rmse_baseline * 100, 2)`% over the
baseline. The step-by-step analysis shows that movie and user biases
account for the largest RMSE reductions, with genre and temporal effects
contributing additional gains. All estimates are regularized with an
optimally tuned $\lambda$ and converge to a stable solution through
iterative updates.

### The Danger of RMSE Delusion

Due to the bias absorption by movie_bias a single RMSE number from a
standard random split can be misleading if interpreted without
considering the deployment task. For a static catalogue (predicting
ratings for the same set of movies), dropping negligible biases like
genre (0.0001 improvement) does not hurt accuracy. For a cold‑start
system (recommending new movies), genre becomes critical – but its value
is invisible in the standard table because movie bias already absorbs
genre signal.

Therefore, always ask: *What is the intended use case?* Then design the
evaluation protocol and model accordingly. The step‑wise RMSE table is a
tool, not a verdict. in general the [Step-by-Step RMSE Improvement]
table is a representative of the base the model was been built for and
doesn't mean genre isn't a valuable feature we can use to forecast the
rating perhaps it could be very useful feature if the task change even
if the RMSE scale show it's not

## Limitations

The model is linear and additive, missing potential interaction effects
(e.g., a user who specifically loves action films). Temporal effects
(year, month, week) contribute modestly compared to movie and user
biases.

## Potential Extensions

-   **Deployment ready:** The saved predictor function allows easy
    integration into a production environment (e.g., a Shiny app or API)
    without any R Markdown or training code.

-   Matrix factorisation (SVD/ALS) to capture latent user‑item
    interactions.

-   More granular temporal dynamics (e.g., drifting user biases over
    time).

-   Ensemble with a latent factor model.

------------------------------------------------------------------------

*Report generated using R Markdown. All analysis was performed in R
using the `tidyverse`, `data.table`, `caret`, and `lubridate` packages.
Results are based on pre‑computed outputs from the final model script.*
